Modal Abstractions for Virtualizing Memory Addresses

### 2023-02-27 programlogic.tex
+

  \subsection{Aliasing/Sharing Physical Pages}
  \label{sec:sharingpages}
  We already build the intuition on what a virtual address mapping is composed of: 1. traversing the page tables to reach the page table entry 2. fetching the physical address, and accessing the value mapped. This two fold access-pattern enabling sharing of pages by virtual memory addresses exhibits logically sound representation against virtual memory operations some of which ((e.g. mapping a new page) we explain in detail in Section \ref{}.

One might have already observed that this virtual points-to definition is too strong to specify virtual memory operations, some of which we explain within as part of experiement in Section \ref{}, which  requires reasoning of sharing physical pages across different virtual memory addresses. Although there seems to fractional permissions for fragmental ownership of page-table entries, strong virtual points-to (in Figure \ref{strongvirtualpointsto}) relation doesn’t care about owning those fractional resources, it only cares that something ensures the correct page table walk exists. Therefore we need to weaken the virtual-pointsto to capture the acutal sharing accross the virtual addresses.

One possible approach to is to just abstracting the \textit{valid} page-table walk existence from a each virtual address (e.g. \textsf{va}) to a physical page address, as \textit{per-element} ownership of a logical \textit{page-table-walk-summarization-map}
\[\mathcal{M}\textsf{PTableWalk} \stackrel{\triangle}{=} \textsc{Auth} (\; \mathcal{W}_{64} \;\rightarrow_{\textrm{fin}} \;  (\mathcal{W}_{64} \; \rightarrow_{\textrm{fin}}\;  (\textsc{Frac }, \mathord{+}) \times (\textsc{Agree } \Loc,\mathord{=}) ))\]
provided by a fragmentery ghost resrource ownership of $\mathcal{A}\textsf{bsPTableWalk}$.
\begin{remark}[Exchanging Page-Table-Walk-Summarization Mappings (Tokens) with Physical Page-Table-Walk Pointstos ]
$\mathcal{A}\textsf{PTableWalk}$ is a ghost map resource where fragmental-ownership of an individual key-value pair (per-element) can be handled as a resource. We use this mechanism to separate the virtual-to-physical translation from the physical memory involved: virtual points-to assertions can be defined in terms of fragmental ownership (ownership of an element) of these partial ghost map entry assertions , $\sumwalkabs\vaddr\qfrac\paddr$, plus sharing an invariant ($\mathcal{I}$\textsf{ASpace} in Figure \ref{fig:peraspaceinvariant}) that the current installed page table respects all entries of the mapping, $\ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}}$. From one another perspective, unmapping pages collects the fragmental map kvpairs, e.g. $\sumwalkabs\vaddr\qfrac\paddr$, from collecting the assertions, and then can remove them from the ghost map and update the page tables. Critically, physical full ownership of the page tables lives, $\ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}}$, in the central invariant on the current page table, so some virtual pointsto assertions can refer to memory in those page tables.% This still works with the modality, since that invariant is also semantically a predicate on a page table root.
\end{remark}
\begin{figure*}
\[
\begin{array}{l}
  \textsf{VirtPtsTo}\stackrel{def}{=} \lambda \tlf\Loc.\\
  \ownGhost\gamma{a} 
\end{array}
\]
\caption{Virtual-Pointsto for Sharing Pages}
\todo[inline]{These fractions aren't quite right (though I see you added the variance between levels), I can walk you through in our meeting tomorrow.}
  \label{fig:virtualpointstosharing}
\end{figure*}

\subsection{Address-Space Abstraction}
\label{sec:aspacemodal}
Abstractions we introduce for virtual address spaces are based on two pillars: 1. a global address space invariant asserting the authorative ownership of the ghost map, , summarizing the existence of the page-table walks, and for each physical page address reached by one of these abstract page-table walks we know that there exists concrete physical page-table accesses, i.e. L4-L1 pointsto  2. a modal definition describing the truth of assertions for the resources inside the address space, $[r]P$ indicating that $P$ holds in the virtual address space rooted at $r$.  This is essentially what hybrid logic \ref{} calls a satisfaction operator, which evaluates the truth of a predicate in a named alternative state (here, address space). 
\subsubsection{Global Per-Addressspace Invariant}
\label{sec:peraspaceinvariant}
\begin{figure*}
\[
\begin{array}{l}
  \textsf{ASpaceInv}\stackrel{def}{=} \lambda \tlf\Loc.\\
  \ownGhost\gamma{a}
\end{array}
\]
\caption{Global Address-Space Invarian}
\todo[inline]{These fractions aren't quite right (though I see you added the variance between levels), I can walk you through in our meeting tomorrow.}
  \label{fig:peraspaceinvariant}
\end{figure*}
\subsubsection{Address-Space Modality}
\label{sec:aspacemodalist}
\paragraph{Modal Resource Context}
\label{sec:resourcecontext}
\todo[inline,color=green]{ismail, speak about virtual points-to established under a certain root of an address space}
\paragraph{Interaction}
\label{sec:interaction}
\todo[inline,color=green]{ismail: speak about the interection with ambient logic}
\subsection{Considering Hoare Doubles in the Context of Switching Address-Space}
An important subtlety arises with supporting \lstinline|mov|s into \lstinline|%cr3|. Consider the hypothetical rule:
\begin{mathpar}
\inferrule[Broken]{ }{
  \{P \ast cr3=r_1 \ast r=r_2 \ast [r_2](Q)\}
  \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
  \{[r_1](P) \ast cr3=r_2 \ast r = r_2 \ast Q\}
}
\end{mathpar}
This rule captures the intuitive change of address space in a Hoare triple, rather than double, form. The problem with this is that it interacts quite poorly with the traditional frame rule and the modal flavor of virtual points-to assertions:
\begin{mathpar}
  \inferrule*[right=Frame]{
    \inferrule*[right=Broken]{ }{
    \{\mathsf{emp} \ast cr3=r_1 \ast r=r_2 \ast [r_2](Q)\}
    \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
    \{[r_1](\mathsf{emp}) \ast cr3=r_1 \ast r=r_2 \ast Q\}
    }
  }{
    \{a\mapsto_\mathsf{v} x \ast \mathsf{emp} \ast cr3=r_1 \ast r=r_2 \ast [r_2](Q)\}
    \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
    \{a\mapsto_\mathsf{v} x \ast [r_1](\mathsf{emp}) \ast cr3=r_1 \ast r=r_2 \ast Q\}
  }
\end{mathpar}
Notice that both the precondition and postcondition assert that $a\mapsto_\mathsf{v} x$ in the current address space, but we have no basis for concluding that address translation is preserved by the change of address space. So this derivation clearly leads to an unsound conclusion. This suggestss that the traditional frame rule and the Hoare triple presentation of the change-of-address-space rule cannot soundly coexist in the same system.
The heart of the problem is that while updating \lstinline|cr3| is \emph{physically} local, it globally changes the interpretation of virtual addresses. So it is simply unsound to frame around \lstinline|cr3| updates.

>>>>

  \subsection{Aliasing/Sharing Physical Pages}
  \label{sec:sharingpages}
  We already build the intuition on what a virtual address mapping is composed of: 1. traversing the page tables to reach the page table entry 2. obtaining the physical address, and accessing the value mapped.

One might have already observed that the virtual points-to definition shown in Figure \ref{fig:strongvirtualpointsto} is too strong to specify virtual memory operations updating a page table entry. Although there seems to fractional permissions for fragmental ownership of page-table entries, strong virtual points-to (in Figure \ref{strongvirtualpointsto}) relation doesn’t care about owning those fractional resources, it only cares that something ensures the correct page table walk exists. Any virtual memory operations such as mapping a new page which updates an page table entry would violate the soundness of other virtual-pointsto relations in the existence of page-sharing across the virtual addresses in the same address space.

One possible approach to is to just abstracting the \textit{valid} page-table walk existence from a each virtual address (e.g. \textsf{va}) to a physical page address, as \textit{per-element} ownership of a logical \textit{page-table-walk-summarization-map}
\[\mathcal{A}\textsf{bsPTableWalk} \stackrel{\triangle}{=} \textsc{Auth} (\; \mathcal{W}_{64} \;\rightarrow_{\textrm{fin}} \;  (\mathcal{W}_{64} \; \rightarrow_{\textrm{fin}}\;  (\textsc{Frac }, \mathord{+}) \times (\textsc{Agree } \Loc,\mathord{=}) ))\]
provided by a fragmentery ghost resrource ownership of $\mathcal{A}\textsf{bsPTableWalk}$.
\begin{remark}[Exchanging Page-Table-Walk-Summarization Mappings (Tokens) with Physical Page-Table-Walk Pointstos ]
$\mathcal{A}\textsf{PTableWalk}$ is a ghost map resource where fragmental-ownership of an individual key-value pair (per-element) can be handled as a resource. We use this mechanism to separate the virtual-to-physical translation from the physical memory involved: virtual points-to assertions can be defined in terms of fragmental ownership (ownership of an element) of these partial ghost map entry assertions , $\sumwalkabs\vaddr\qfrac\paddr$, plus sharing an invariant ($\mathcal{I}$\textsf{ASpace} in Figure \ref{fig:peraspaceinvariant}) that the current installed page table respects all entries of the mapping, $\ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}}$. From one another perspective, unmapping pages collects the fragmental map kvpairs, e.g. $\sumwalkabs\vaddr\qfrac\paddr$, from collecting the assertions, and then can remove them from the ghost map and update the page tables. Critically, physical full ownership of the page tables lives, $\ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}}$, in the central invariant on the current page table, so some virtual pointsto assertions can refer to memory in those page tables.% This still works with the modality, since that invariant is also semantically a predicate on a page table root.
\end{remark}
Then, our virtual memory addressing would be the fragmental ownership of the abstract page-table-walk mapping with the physical page address mapping as shown in Figure \ref{fig:virtualpointstosharing}. 
\begin{figure*}
\[
\begin{array}{l}
  \textsf{VPointsTo}(\vaddr \;, \vpage)\stackrel{\triangle}{=} 
  \exists \entryo \ldotp 
  \sumwalkabs\vaddr\qfrac\paddr \star 
   \nfpointsto{\mask\vaddr\ft\entryo}{\mask\vaddr\tw\entryo}\vpage\qfracone\naddr
\end{array}
\]
\caption{Virtual-Pointsto for Sharing Pages}
  \label{fig:virtualpointstosharing}
\end{figure*}

\subsection{Address-Space Abstraction}
\label{sec:aspacemodal}
Abstractions we introduce for virtual address spaces are based on two pillars:
\begin{enumerate}
\item a global address space invariant asserting the authorative ownership of the ghost map, , summarizing the existence of the page-table walks, and for each physical page address reached by one of these abstract page-table walks we know that there exists concrete physical page-table accesses, i.e. L4-L1 pointsto
  \item  a modal definition describing the truth of assertions for the resources inside the address space, $[r]P$ indicating that $P$ holds in the virtual address space rooted at $r$.  This is essentially what hybrid logic \ref{} calls a satisfaction operator, which evaluates the truth of a predicate in a named alternative state (here, address space). 
\end{enumerate}
    \subsubsection{Global Per-Addressspace Invariant}

\label{sec:peraspaceinvariant}
\begin{figure*}
\[
\begin{array}{l}
  \mathcal{I}\textsf{ASpace} \stackrel{def}{=} \lambda \tlf\Loc.\\
  \ownGhost\gamma{a}
\end{array}
\]
\caption{Global Address-Space Invarian}
\todo[inline]{These fractions aren't quite right (though I see you added the variance between levels), I can walk you through in our meeting tomorrow.}
  \label{fig:peraspaceinvariant}
\end{figure*}
\subsubsection{Address-Space Modality}
\label{sec:aspacemodalist}
\paragraph{Modal Resource Context}
\label{sec:resourcecontext}
\todo[inline,color=green]{ismail, speak about virtual points-to established under a certain root of an address space}
\paragraph{Interaction}
\label{sec:interaction}
\todo[inline,color=green]{ismail: speak about the interection with ambient logic}
\subsection{Considering Hoare Doubles in the Context of Switching Address-Space}
An important subtlety arises with supporting \lstinline|mov|s into \lstinline|%cr3|. Consider the hypothetical rule:
\begin{mathpar}
\inferrule[Broken]{ }{
  \{P \ast cr3=r_1 \ast r=r_2 \ast [r_2](Q)\}
  \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
  \{[r_1](P) \ast cr3=r_2 \ast r = r_2 \ast Q\}
}
\end{mathpar}
This rule captures the intuitive change of address space in a Hoare triple, rather than double, form. The problem with this is that it interacts quite poorly with the traditional frame rule and the modal flavor of virtual points-to assertions:
\begin{mathpar}
  \inferrule*[right=Frame]{
    \inferrule*[right=Broken]{ }{
    \{\mathsf{emp} \ast cr3=r_1 \ast r=r_2 \ast [r_2](Q)\}
    \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
    \{[r_1](\mathsf{emp}) \ast cr3=r_1 \ast r=r_2 \ast Q\}
    }
  }{
    \{a\mapsto_\mathsf{v} x \ast \mathsf{emp} \ast cr3=r_1 \ast r=r_2 \ast [r_2](Q)\}
    \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
    \{a\mapsto_\mathsf{v} x \ast [r_1](\mathsf{emp}) \ast cr3=r_1 \ast r=r_2 \ast Q\}
  }
\end{mathpar}
Notice that both the precondition and postcondition assert that $a\mapsto_\mathsf{v} x$ in the current address space, but we have no basis for concluding that address translation is preserved by the change of address space. So this derivation clearly leads to an unsound conclusion. This suggestss that the traditional frame rule and the Hoare triple presentation of the change-of-address-space rule cannot soundly coexist in the same system.
The heart of the problem is that while updating \lstinline|cr3| is \emph{physically} local, it globally changes the interpretation of virtual addresses. So it is simply unsound to frame around \lstinline|cr3| updates.
